---
title: "ASML Evaluation ex2"
author: "Sophie Gallet"
date: "1/20/2020"
output: html_document
---

# Modeling Real Estate Pricing

## Introduction

```{r load, message = FALSE}
load("ames_train.Rdata")
```

```{r packages, message = FALSE}
library(statsr)
library(dplyr)
library(BAS)
library(kableExtra)
library(ggplot2)
library(MASS)
library(olsrr)

# Helper function
get_rmse <- function(mod, data){
  predictions <- exp(predict(mod, data))
  residuals <- data$price - predictions
  rmse <- sqrt(mean(residuals**2))
}
```

* * *

## Exploratory Data Analysis (EDA)

```{r}
dim(ames_train)
summary(ames_train)
```

```{r data-manipulation}
#Keep only observations sold under normal conditions
ames_train <- ames_train %>%
  filter(Sale.Condition == "Normal")

#Recode as a separate category NA values in Bsmt.Qual, BsmtFin.Type.1, BsmtFin.Type.2
levels(ames_train$Bsmt.Qual) <- c(levels(ames_train$Bsmt.Qual),"no_bsmt")
ames_train$Bsmt.Qual[is.na(ames_train$Bsmt.Qual)] <- "no_bsmt"

levels(ames_train$BsmtFin.Type.1) <- c(levels(ames_train$BsmtFin.Type.1),"no_bsmt")
ames_train$BsmtFin.Type.1[is.na(ames_train$BsmtFin.Type.1)] <- "no_bsmt"

levels(ames_train$BsmtFin.Type.2) <- c(levels(ames_train$BsmtFin.Type.2),"no_bsmt")
ames_train$BsmtFin.Type.2[is.na(ames_train$BsmtFin.Type.2)] <- "no_bsmt"

#Recode as a separate category NA values in Garage.Qual
levels(ames_train$Garage.Qual) <- c(levels(ames_train$Garage.Qual),"no_garage")
ames_train$Garage.Qual[is.na(ames_train$Garage.Qual)] <- "no_garage"
```


```{r log_area}
ames_train <- ames_train %>%
  mutate(log_area = log(area),
         log_price = log(price),
         log_Lot.Area = log(Lot.Area))

ames_train %>%
  ggplot(aes(log_area,log_price, col = log_price)) + 
  labs(title = "Log(Area) & Log(price)", x = "log(area)", y = "log(price)") + 
  geom_smooth(method='lm',formula = y ~ x) + 
  geom_point() +
  theme_classic() + 
  guides(col = FALSE) 
```
```{r overall_qual}
ames_train %>%
  group_by(factor(Overall.Qual)) %>%
  mutate(median.price = median(price)) %>%
  ggplot(aes(factor(Overall.Qual), price, fill = median.price)) + 
  geom_boxplot() + 
  theme_classic() +
  labs(title = "Overall Quality & Price", x = "Overall Quality", y = "Price") +
  guides(fill = FALSE) + 
  scale_fill_gradient( low = "#D8F9FF", high = "#00A190") 
```

```{r neighborhood}
ames_train %>%
  group_by(Neighborhood) %>%
  mutate(median.price = median(price)) %>%
  ggplot(aes(Neighborhood, price, fill = median.price)) + 
  labs(title = "Neighborhood & Price") + 
  geom_boxplot() +
  theme_classic() + 
  guides(fill = FALSE) + 
  coord_flip() +
  scale_fill_gradient( low = "#D8F9FF", high = "#00A190") 
```


* * *

## Initial model

```{r fit_model}
mod1 <- lm(log_price ~ log_area + log_Lot.Area + Overall.Qual + Neighborhood + Bedroom.AbvGr + Year.Built + Garage.Cars + Garage.Qual + X1st.Flr.SF + Bsmt.Qual,
                   data = ames_train)
summary(mod1)
```

### Variable Selection


```{r}
variable_selection <-ols_step_best_subset(mod1)
variable_selection
```

```{r}
plot(variable_selection)
```


```{r}
mod2 <- lm(log_price ~ log_area + log_Lot.Area + Overall.Qual + Neighborhood + Year.Built + Garage.Qual + X1st.Flr.SF + Bsmt.Qual,
                   data = ames_train)
data2 <- subset(ames_train, select = c("log_price", "log_area", "log_Lot.Area", "Overall.Qual","Neighborhood", "Year.Built", "Garage.Qual", "X1st.Flr.SF", "Bsmt.Qual"))
summary(mod2)
```


### Simple Model Residuals


```{r}
hist(mod2$residuals)
plot(mod2)
ggplot(data = mod2, aes(x = .fitted, y = .resid)) +
  geom_point(col = ifelse(abs(mod2$residuals)> mean(mod2$residuals) +
3*sd(mod2$residuals), "red", "black"), 
            size = ifelse(abs(mod2$residuals)> mean(mod2$residuals) +
3*sd(mod2$residuals), 2, 1)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals") +
  theme_classic() 
```




```{r}
data2$res2 <- mod2$residuals

data2 %>%
  arrange(desc(abs(mod2$residuals))) %>%
  head(10)

summary(data2$Year.Built)
```


### Initial Model RMSE

```{r init_model_rmse_train}
rmse.initial_train <- get_rmse(mod2, ames_train)
c("Training RMSE ($)" = rmse.initial_train)
```

### Overfitting 

The first chunk of code loads the test data and performs the same manipulations as done for the training data. 

```{r loadtest}
load("ames_test.Rdata")

ames_test <- ames_test %>%
  mutate(log_area = log(area),
         log_price = log(price),
         log_Lot.Area = log(Lot.Area))

#Remove observation from Landmark Neighborhood (no such observations in training data), keep observations with normal sale conditions 
ames_test <- ames_test %>%
  filter(ames_test$Neighborhood != "Landmrk",
         Sale.Condition == "Normal")

#Recode as a separate category NA values in Bsmt.Qual, BsmtFin.Type.1,BsmtFin.Type.2
levels(ames_test$Bsmt.Qual) <- c(levels(ames_test$Bsmt.Qual),"no_bsmt")
ames_test$Bsmt.Qual[is.na(ames_test$Bsmt.Qual)] <- "no_bsmt"

levels(ames_test$BsmtFin.Type.1) <- c(levels(ames_test$BsmtFin.Type.1),"no_bsmt")
ames_test$BsmtFin.Type.1[is.na(ames_test$BsmtFin.Type.1)] <- "no_bsmt"

levels(ames_test$BsmtFin.Type.2) <- c(levels(ames_test$BsmtFin.Type.2),"no_bsmt")
ames_test$BsmtFin.Type.2[is.na(ames_test$BsmtFin.Type.2)] <- "no_bsmt"

#Recode as a separate category NA values in Garage.Qual
levels(ames_test$Garage.Qual)<- c(levels(ames_test$Garage.Qual),"no_garage")
ames_test$Garage.Qual[is.na(ames_test$Garage.Qual)] <- "no_garage"

```


```{r init_model_rmse_test}
rmse.initial_test<- get_rmse(mod2, ames_test)

c("RMSE Test ($)" = rmse.initial_test, "RMSE Train ($)" = rmse.initial_train)
```

* * *

## Final Model

```{r}
#Add interaction terms for garage
ames_train <- ames_train %>%
  mutate(Garage = Garage.Cars * Garage.Area,
         Bath = Full.Bath * Half.Bath) #used in full model, discarded for final model
ames_test <- ames_test %>%
  mutate(Garage = Garage.Cars * Garage.Area,
         Bath = Full.Bath * Half.Bath) #used in full model, discarded for final model

# "Big" model (20 variables)
big.model <- lm(log_price ~ log_area + log_Lot.Area + BsmtFin.Type.1 + log(Total.Bsmt.SF+1) + log(BsmtFin.SF.1+1) + Bsmt.Qual + Overall.Qual + Neighborhood + Year.Built + Year.Remod.Add + Bath + Garage.Qual + Garage + Exter.Qual + Bldg.Type + Central.Air, 
                  data = ames_train)

#Final model & summary
final.model <- lm(log_price ~ log_area + log_Lot.Area + BsmtFin.Type.1 + log(Total.Bsmt.SF + 
    1)  + Overall.Qual + Neighborhood + Year.Built + Year.Remod.Add + Garage + Exter.Qual + 
    Central.Air, 
                  data = ames_train)
summary(final.model)
```


### Transformation

```{r model_assess}
par(mfrow = c(2,2))
hist(ames_train$price, main = "Histogram of price", xlab = "price")
hist(ames_train$area, breaks = 20, main = "Histogram of area", xlab = "area")
hist(ames_train$Lot.Area, breaks = 50, main = "Histogram of Lot.Area", xlab = "Lot.Area")
hist(ames_train$Total.Bsmt.SF, main = "Histogram of Total.Bsmt.SF", xlab = "Total.Bsmt.SF")
```


### Variable Interaction

```{r model_inter}
plot(ames_train$Garage.Cars, ames_train$Garage.Area, xlab = "Garage.Cars", ylab = "Garage.Area")
```

* * *

### Variable Selection


```{r model_select}
model.aic <- stepAIC(big.model, trace=F, direction="both")
#summary(big.model)
summary(model.aic)
```

### Model Testing

```{r model_testing}

# AIC model - train data
rmse.aic_train <- get_rmse(model.aic, ames_train)
# AIC model - test data
rmse.aic_test <- get_rmse(model.aic, ames_test)

# Final model - train data
rmse.final_train <- get_rmse(final.model, ames_train)
# Final model - test data
rmse.final_test <- get_rmse(final.model, ames_test)


c("RMSE Summary")
c("AIC Test ($)" = rmse.aic_test, "AIC Train ($)" = rmse.aic_train, "% difference" = round((1 - rmse.aic_train/rmse.aic_test)*100,2))
c("Final Test ($)" = rmse.final_test, "Final Train ($)" = rmse.final_train, "% difference" = round((1 - rmse.final_train/rmse.final_test)*100,2))

#Coverage Probabilities
# Predict prices - Test
predict.test <- exp(predict(final.model, ames_test, interval = "prediction"))

# Calculate proportion of observations that fall within prediction intervals - Test
coverage.prob.test <- mean(ames_test$price > predict.test[,"lwr"] &
                            ames_test$price < predict.test[,"upr"])

# Predict prices - Train
predict.train <- exp(predict(final.model, ames_train, interval = "prediction"))

# Calculate proportion of observations that fall within prediction intervals - Train
coverage.prob.train <- mean(ames_train$price > predict.train[,"lwr"] &
                            ames_train$price < predict.train[,"upr"])

c("Coverage Probability test" = coverage.prob.test, "Coverage Probability train" = coverage.prob.train)
```

* * *

## Final Model Assessment

### Final Model Residuals

```{r final_residual_plot}
ggplot(data = final.model, aes(x = .fitted, y = .resid)) +
  geom_point(col = ifelse(abs(final.model$residuals)> mean(final.model$residuals) +
3*sd(final.model$residuals), "red", "black"), 
            size = ifelse(abs(final.model$residuals)> mean(final.model$residuals) +
3*sd(final.model$residuals), 2, 1)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals") +
  theme_classic() 
```

### Final Model RMSE

```{r final_model_testing}
#Display RMSE (already calculated
c("RMSE Test ($)" = rmse.final_test, "RMSE Train ($)" = rmse.final_train)
```

### Final Model Validation

```{r loadvalidation, message = FALSE}
load("ames_validation.Rdata")
```


```{r model_validate}
#Data Manipulation on Validation data
#Keep observations with normal sale conditions 
ames_validation <- ames_validation %>%
  filter(Sale.Condition == "Normal") %>%
  mutate(log_area = log(area),
         log_price = log(price),
         log_Lot.Area = log(Lot.Area))

#Recode as a separate category NA values in BsmtFin.Type.1
levels(ames_validation$BsmtFin.Type.1) <- c(levels(ames_validation$BsmtFin.Type.1),"no_bsmt")
ames_validation$BsmtFin.Type.1[is.na(ames_validation$BsmtFin.Type.1)] <- "no_bsmt"

#Create Garage tnteraction term
ames_validation <- ames_validation %>%
  mutate(Garage = Garage.Cars * Garage.Area)



# RMSE Validation data
rmse.final_val <- get_rmse(final.model, ames_validation)
c("RMSE Val" = rmse.final_val, "RMSE Test" = rmse.final_test, "RMSE Train" = rmse.final_train) 


#Coverage Probability
# Predict prices
predict.val <- exp(predict(final.model, ames_validation, interval = "prediction"))

# Calculate proportion of observations that fall within prediction intervals
coverage.prob.val <- mean(ames_validation$price > predict.val[,"lwr"] &
                            ames_validation$price < predict.val[,"upr"])
c("Coverage Probability" = coverage.prob.val)
```

* * *
